---
title: "R-programming of ALL(B-cell)"
author: "Stephen Omitogun"
date: "February 15, 2016"
output: pdf_document
---

The library containing the ALL(acute lymphoblastic leukemia) data was downloaded from bioconductor.org through some commands.
biocLite("ALL")
biocLite("genefilter")

To open the ALL dataset, the following commands are:
```{r}
library(Biobase)
library(ALL)
data(ALL)
```

##Investigating the ALL dataset##

```{r, echo=TRUE, results='markup'}
class(ALL)
```

```{r, echo=TRUE, results='markup'}
ALL1 <- data.frame(ALL)
```

```{r, echo=TRUE, results='markup'}
dim(ALL)
```

```{r, echo=TRUE, tidy=TRUE, results='asis'}
str(ALL)
```


###Information about assay and sample data###


```{r, echo=TRUE, tidy=TRUE}
phenoData(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
varLabels(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
varMetadata(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
featureNames(ALL)[1:128]
```
The above results are the feature names.

#####The patient id#####

```{r, echo=TRUE, tidy=TRUE}
sampleNames(ALL)[1:20]
```

```{r, echo=TRUE, tidy=TRUE}
expressionData <- exprs(ALL)
str(expressionData)
```
We can look at the expression rate at which individual patient gene were expressed. Below two feature names and expresion levels for each patient within the two features are computated.

```{r, echo=TRUE, tidy=TRUE}
gene.expression <- ALL[c("1000_at", "1001_at"),]
exprs(gene.expression)
```

```{r, echo=TRUE, tidy=TRUE}
pData(ALL)[1:15,]
```
######ALL phenotype data######


**Subsetting B-cell patients from the ALL dataset**

```{r, echo=TRUE, tidy=TRUE}
ALL$BT

anyB <- grep("^B", ALL$BT)
b.cell <- ALL[, anyB]

```

```{r, echo=TRUE, tidy=TRUE}

exprsData = exprs(b.cell)
dim(exprsData)

pData = pData(b.cell)
dim(pData)
head(pData)

featureData = featureData(b.cell)
dim(featureData)

table(pData$sex, pData$age)

table(pData$mol.biol, pData$BT)

table(pData$age, pData$BT)

summary(exprsData)

```

**Reducing genes based on variability**

```{r, echo=TRUE, tidy=TRUE, results='hide', warning=FALSE}
tropical= c('darkorange', 'dodgerblue', 'hotpink', 'limegreen','yellow')

palette(tropical)
par(pch=19)
library(genefilter)
library(RColorBrewer)
library(devtools)
library(RSkittleBrewer)
library(gplots)
library(dplyr)
library(AnnotationDbi)

```


```{r}
##Boxplot applied to entire expression matrix
boxplot(log2(exprsData+1),col=2,range=0)

par(mfrow=c(1,2)) ##<- setting up parameter for plots. 
hist(log2(exprsData[,1]+1), col=2)

hist(log2(exprsData[,2]+1), col=2)

par(mfrow=c(1,1))

plot(density(log2(exprsData[,2]+1)),col=2)

###The lines command will allow to overlay another plot on top of the plot previously

lines(density(log2(exprsData[,2]+1)),col=3)

qqplot(log2(exprsData[,1]+1), log2(exprsData[,2]+1), col=3)

### I can also use the qqplot to see if the samples are consistant. qqplot is making one dot for every quantile of this two distribution.

abline(c(0,1)) ###<- creates a 45 degree line.

mm = log2(exprsData[,1]+1) - log2(exprsData[,2]+1)

aa = log2(exprsData[,1]+1) + log2(exprsData[,2]+1)

plot(aa, mm, col=2)

lowgenes = rowMeans(exprsData) < 8.5
table(lowgenes)
eData_filt = filter(as.data.frame(exprsData), !lowgenes)
dim(eData_filt)

lowgenes2 = rowMedians(as.matrix(exprsData)) < 8.5

table(lowgenes2)
table(lowgenes2, lowgenes)

exprsData_df = data.frame(exprsData)

eData_filt1 = filter(exprsData_df, !lowgenes2)

eData_filt1 = log2(eData_filt1 +1) ###<- To remove aswell the undefined values.
dim(eData_filt1)

hist(log2(eData_filt1[,1]+1), col=2)

boxplot(as.matrix(log2(eData_filt1+8.5)),col=2)
```

```{r}
boxplot(colSums(eData_filt1) ~ pData$sex, col=2) ###<- comparing the gene expression between male and female using a box plot.

boxplot(colSums(eData_filt1) ~ pData$sex)

points(colSums(eData_filt1) ~ jitter(as.numeric(pData$sex)), col=as.numeric(pData$sex), pch=19) ###<- adding points to the boxplot exploratory data to analyse the comparison.

plot(colSums(eData_filt1) ~ pData$age)

boxplot(colSums(eData_filt1) ~ pData$BT)

points(colSums(eData_filt1) ~ jitter(as.numeric(pData$BT)), col=as.numeric(pData$BT), pch=19)
eData_filt1_matrix = as.matrix(eData_filt1)

boxplot(colSums(eData_filt1) ~ pData$`t(4;11)`)

boxplot(colSums(eData_filt1) ~ pData$`t(9;22)`)

points(colSums(eData_filt1) ~ jitter(as.numeric(pData$`t(9;22)`)), col=as.numeric(pData$`t(9;22)`), pch=19)

###<- 
boxplot(colSums(eData_filt1) ~ pData$mol.biol)
points(colSums(eData_filt1) ~ jitter(as.numeric(pData$mol.biol)), col=as.numeric(pData$mol.biol), pch=19)
plot(colSums(eData_filt1) ~ pData$mol.biol + pData$age)
boxplot(colSums(eData_filt1) ~ pData$mol.biol + pData$sex)

```


```{r}
eData_filt1_matrix = as.matrix(eData_filt1)

colramp = colorRampPalette(c(3, "white",2))(9)

heatmap(eData_filt1_matrix, col=colramp)

heatmap(eData_filt1_matrix, col=colramp, Rowv = NA, Colv = NA)

heatmap.2(eData_filt1_matrix, col = colramp, Rowv = NA, Colv = NA, dendrogram = "none", scale = "row", trace = "none")

```

***Clustering***

```{r}
library(dendextend)

dist1 = dist(t(eData_filt1))

heatmap(as.matrix(dist1), col=colramp, Colv = NA, Rowv = NA)

hclust1 = hclust(dist1)

plot(hclust1)
plot(hclust1, hang=-1)

dend = as.dendrogram(hclust1)

dend = color_labels(hclust1,4,1:14)

dend = color_labels(hclust1,4,1:4)

dend = color_labels(hclust1,4,1:4)

plot(dend)

dend = color_labels(hclust1,3,1:3)

plot(dend)

kmeans = kmeans(eData_filt1, centers = 3)
names(kmeans)

matplot(t(kmeans$centers))

matplot(t(kmeans$centers), col=1:3, type = "l", lwd=3)

table(kmeans$cluster)

table(kmeans$cluster)

newdata = as.matrix(eData_filt1)[order(kmeans$cluster),]

heatmap(newdata, col = colramp, Colv = NA, Rowv = NA)
```
