---
title: "R-programming of ALL(B-cell)"
author: "Stephen Omitogun"
date: "February 15, 2016"
output: pdf_document
---

The library containing the ALL(acute lymphoblastic leukemia) data was downloaded from bioconductor.org through some commands.
biocLite("ALL")
biocLite("genefilter")

To open the ALL dataset, the following commands are:
```{r}
library(Biobase)
library(ALL)
data(ALL)
```

##Investigating the ALL dataset##

```{r, echo=TRUE, results='markup'}
class(ALL)
```

```{r, echo=TRUE, results='markup'}
ALL1 <- data.frame(ALL)
```

```{r, echo=TRUE, results='markup'}
dim(ALL)
```

```{r, echo=TRUE, tidy=TRUE, results='asis'}
str(ALL)
```


###Information about assay and sample data###


```{r, echo=TRUE, tidy=TRUE}
phenoData(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
varLabels(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
varMetadata(ALL)
```

```{r, echo=TRUE, tidy=TRUE}
featureNames(ALL)[1:128]
```
The above results are the feature names.

#####The patient id#####

```{r, echo=TRUE, tidy=TRUE}
sampleNames(ALL)[1:20]
```

```{r, echo=TRUE, tidy=TRUE}
expressionData <- exprs(ALL)
str(expressionData)
```
We can look at the expression rate at which individual patient gene were expressed. Below two feature names and expresion levels for each patient within the two features are computated.

```{r, echo=TRUE, tidy=TRUE}
gene.expression <- ALL[c("1000_at", "1001_at"),]
exprs(gene.expression)
```

```{r, echo=TRUE, tidy=TRUE}
pData(ALL)[1:15,]
```
######ALL phenotype data######

####Converting ALL data into a data frame####

```{r, echo=TRUE, tidy=TRUE}
ALL1 <- data.frame(ALL)
length(colnames(ALL1))
head(colnames(ALL1))
tail(colnames(ALL1))
```

**Subsetting B-cell patients from the ALL dataset**

```{r, echo=TRUE, tidy=TRUE}
ALL$BT

anyB <- grep("^B", ALL$BT)
b.cell <- ALL[, anyB]

```

```{r, echo=TRUE, tidy=TRUE}
table(pData(b.cell)$sex)

table(pData(b.cell)$mol.biol)

table(pData(b.cell)$age)

table(pData(b.cell)$kinet)

table(pData(b.cell)$cyto.normal)

table(pData(b.cell)$citog)
```

**Reducing genes based on variability**

```{r, echo=TRUE, tidy=TRUE, }
library(genefilter)
library(RColorBrewer)
```


```{r}
B.cell1 = 2^exprs(b.cell)
mean_gene = rowMeans(B.cell1)
sd_gene = rowSds(B.cell1)
cv = sd_gene/mean_gene
blues.ramp = colorRampPalette(brewer.pal(9, "Reds")[-1])
dCol = densCols(log(mean_gene), log(sd_gene), colramp = blues.ramp)
par(mfrow=c(2,2))

plot(mean_gene, sd_gene, log='xy', col=dCol, pch = 16, cex=0.1)
abline(v=100, lwd=3, col='blue')

hist(log(cv), main = NA)
abline(v=log(0.4), lwd=3, col='blue')

B.genes = filterfun(cv(0.4,10))
B.gene.fil = genefilter(B.cell1, B.genes)
B.cell2 = log2(B.cell1[B.gene.fil,])

##correlation of first 5 reduced genes starting with x01005

B.genes = filterfun(cv(0.4,10))
B.gene.fil = genefilter(B.cell1, B.genes)
B.cell2 = log2(B.cell1[B.gene.fil,])
B.cell2_df = data.frame(B.cell2)
B.cell2_df_vars = B.cell2_df[, c(1:5)]
mar.orig = par()$mar
par(mar=c(1,1,1,1))
pairs(B.cell2_df_vars,panel = panel.smooth, col= 9 + B.cell2_df$X01005)
```

```{r}

BT = which(as.character(ALL$BT) %in% c("B", "T"))
Bcell = ALL[, intersect(anyB, BT)]
Bcell$BT = factor(Bcell$BT)
str(Bcell$BT)


B.expressionData = exprs(Bcell)
str(B.expressionData)

B.sds = rowSds(B.expressionData)

hist(B.sds, breaks=50, col="red", xlab="standard deviation of the probes")

B.shorthen = shorth(B.sds)
B.shorthen
abline(v=B.shorthen, col="blue", lwd=3, lty=2)

Bcellfiltered = Bcell[B.sds>=B.shorthen]

dim(exprs(Bcellfiltered))

B.sds_new = rowSds(exprs(Bcellfiltered))

hist(B.sds_new, breaks=50, col="blue", xlab = "new distribution of standard deviation")

B.shorthen_new = shorth(B.sds_new)

B.shorthen_new
abline(v=B.shorthen_new, col="blue", lwd=3, lty=2)

```


Correlation of the first 5 furthered reduced genes 
```{r}
Bcellfiltered_df = data.frame(Bcellfiltered)
Bcellfiltered_vars = Bcellfiltered_df[, c(1:5)]
mar.orig = par()$mar
par(mar = c(1,1,1,1))
pairs(Bcellfiltered_vars, panel = panel.smooth, col= 9 + Bcellfiltered_df$X1003_s_at)

par(mar = rep(2,4))
plot(Bcellfiltered_df$X1003_s_at)
```

```{r}
summary(lm(Bcellfiltered_df$X1003_s_at~Bcellfiltered_df$X1004_at))
Bcell_fit = lm(Bcellfiltered_df$X101_at~Bcellfiltered_df$X1004_at)
anova(Bcell_fit)
par(mar = rep(2,4))
plot(Bcell_fit)
```